---
title: "New York's Streetside Casualties"
subtitle: "An explanatory analysis of NYC car accidents"
format: pdf
geometry: margin=1in
fontsize: 11pt
execute: 
  warning: false
  execute: false
  echo: false
---

# Introduction

## Project Motivation

Citizens of large cities all over America suffer injury and death in motor 
vehicle crashes. In New York, motor vehicle accidents are among the top five 
reasons for hospitalizations statewide (Department of Health, 2023). 
We hope that motivated policymakers in NYC and other metropolitan areas could 
use our explanatory models to craft traffic policy, shift police resources, 
better target traffic citations, and turn our insights into potentially 
lifesaving urban development and planning.

## Dataset

Our dataset is a random sampling of 10,000 motor vehicle collisions (MVCs) out 
of the 2 million MVCs publicly released by the New York City Police
Department (NYPD), spanning all five boroughs of New York City from July 1st, 
2012 through April 24th, 2023. The police report from which individual MVC 
observations in our dataset hail (MV104-AN) is required to be filled out for 
any MVC in which someone is injured or killed, or which result in at least 
\$1,000 of total property damage.

We created a binary response variable corresponding to whether or not an MVC
resulted in a *casualty* (defined as a fatality *or* an injury), and additional
variables corresponding to type of vehicles involved (commercial, motorcycle,
etc.) and factors that contributed to the MVC (aggressive driving, impairment,
etc.). A full data dictionary is available in the appendix.

Our primary research question is determining whether characteristics of an
MVC's timing, contributing factors, and involved vehicles might be associated
with differential odds of casualty.

```{r}
#| label: import-libs
library(tidyverse)
library(Stat2Data)
library(tidymodels)
library(lubridate)
library(flextable)
```

```{r}
#| label: read-data
crashes <- read_csv("data/crashes_10k.csv")
```

## Data Cleaning

For our analysis, we selected only MVCs between two or fewer motorists due to
high levels of missingness in crashes with greater than two motorists. We then
re-categorized the hundreds of vehicle types and contributing factors into 
fewer, larger bins based on common characteristics, and created categorical
variables corresponding to time of day and whether the crash took place during
a weekend.

```{r}
#| label: clean-data
crashes <- crashes |>
  # remove accidents involving greater than 2 vehicles
  filter(is.na(contributing_factor_vehicle_3), is.na(vehicle_type_code_3)) |>
  
  dplyr::select(!c(vehicle_type_code_3, vehicle_type_code_4, vehicle_type_code_5,
            contributing_factor_vehicle_3, contributing_factor_vehicle_4,
            contributing_factor_vehicle_5)) |>
  
  mutate(
    # add combined casualty column (injuries + fatalities)
    num_casualties = number_of_persons_killed + number_of_persons_injured,
    
    # add time of day categories
    time_day = case_when(
      hms(crash_time) > hm("5:00") & hms(crash_time) <= hm("12:00") ~
        "morning",
      hms(crash_time) > hm("12:00") & hms(crash_time) <= hm("17:00") ~
        "afternoon",
      hms(crash_time) > hm("17:00") & hms(crash_time) <= hm("21:00") ~
        "evening",
      hms(crash_time) > hm("21:00") | hms(crash_time) <= hm("5:00") ~
        "night"
    ) %>% 
      factor(levels = c("morning", "afternoon", "evening", "night")),
    
    # create binary variables for injury, fatality, and casualty
    has_injury = number_of_persons_injured > 0,
    has_fatality = number_of_persons_killed >0,
    has_casualty = has_injury | has_fatality,
    # add ordinal column, `severity`, for no casualties v. injuries v. fatalities
    severity = case_when(
      has_fatality ~ "fatal",
      has_injury ~ "injury",
      T ~ "no casualties"
    ),
    severity = factor(severity, levels = c("no casualties", "injury", "fatal")),
    
    # add Julian date column
    crash_date = as.Date(crash_date, format = "%m/%d/%Y"),
    crash_day = weekdays(crash_date, abbreviate = F),
    yday = yday(crash_date),
    
    # add column for weekend or weekday
    weekend_weekday = if_else(crash_day %in% c("Saturday", "Sunday"), 
                              "Weekend", "Weekday"),
    
    # make `crash_day` a factor rather than a character
    crash_day = as.factor(crash_day),
    
    # make `weekend_weekday` a factor rather than a character
    weekend_weekday = as.factor(weekend_weekday)

  )

# consolidate contributing factors
crashes <- crashes |>
  mutate(vehicle_type_code_2 = if_else(
    !is.na(contributing_factor_vehicle_2) & is.na(vehicle_type_code_2),
    "unk",
    vehicle_type_code_2
  ))

crashes <- crashes |> drop_na(vehicle_type_code_1)

ped_bike <- "Pedestrian/Bicyclist/Other Pedestrian Error/Confusion"
crashes <- crashes %>%
  mutate(across(starts_with("contributing_factor_vehicle"),
                ~ case_when(
                  . %in% c("Alcohol Involvement", "Drugs (Illegal)",
                           "Prescription Medication",
                           "Driver Inattention/Distraction","Fatigued/Drowsy",
                           "Lost Consciousness", "Other Electronic Device",
                           "Outside Car Distraction", "Passenger Distraction")
                           ~ "Impairment/Distraction/Fatigue",
                  . %in% c("Following Too Closely", "Passing Too Closely",
                           "Unsafe Lane Changing",
                           "Unsafe Speed",
                           "Backing Unsafely") ~ "Aggressive/Reckless Driving",
                  . %in% c("Failure to Yield Right-of-Way",
                           ped_bike,
                           "Reaction to Other Uninvolved Vehicle",
                           "Reaction to Uninvolved Vehicle",
                           "Turning Improperly")
                           ~ "Failure to Obey Traffic Signs/Signals/Rules",
                  . %in% c("Lane Marking Improper/Inadequate",
                           "Obstruction/Debris", "Other Vehicular",
                           "Oversized Vehicle", "Pavement Defective",
                           "Pavement Slippery",
                           "Traffic Control Device Improper/Non-Working",
                           "Traffic Control Disregarded",
                           "View Obstructed/Limited")
                           ~ "Performance-unrelated Technical/Mechanical Factors",
                  TRUE ~ "Other/Unknown"
                )
  )) |>
  rename(factor1 = contributing_factor_vehicle_1, 
         factor2 = contributing_factor_vehicle_2)

# consolidate vehicle type
crashes <- crashes %>%
  mutate(across(starts_with("vehicle_type_code_1"),
                ~ case_when(
                  . %in% c("2 dr sedan", "3-Door", "4 dr sedan", "4dsd",
                        "Convertible", "Sedan", "SEDONA",
                        "Station Wagon/Sport Utility Vehicle",
                        "SPORT UTILITY / STATION WAGON", "LIMO",
                        "LIVERY VEHICLE") ~ "Passenger vehicles",
                  . %in% c("AMBU", "AMBUL", "Ambulance", "AMBULANCE",
                        "AMBULETTE", "Armored Truck", "Beverage Truck",
                        "Box Truck", "Bulk Agriculture", "Bus", "BUS",
                        "Carry All", "Chassis Cab", "DELIV", "DELV",
                        "Dump", "FDNY", "Fire", "FIRE TRUCK", "FIRETRUCK",
                        "Flat Bed", "Flat Rack", "Garbage or Refuse",
                        "Hopper", "SMALL COM VEH(4 TIRES)",
                        "LARGE COM VEH(6 OR MORE TIRES)",
                        "Mack Truck",
                        "PICK-UP TRUCK", "PK", "Refrigerated Van", "schoo",
                        "SCHOO", "Snow Plow", "Tow Truck",
                        "Tow Truck / Wrecker", "TRACT",
                        "Tractor Truck Diesel", "Tractor Truck Gasoline",
                        "TRAIL", "TRAILER", "TRALI", "TRUCK",
                        "USPS", "UTIL", "Van", "VAN",
                        "VAN TRUCK") ~ "Commercial vehicles",
                  . %in% c("MOTOR SCOO", "Motorcycle",
                        "MOTORCYCLE", "Motorscooter", "SCOOTER", "Moped") ~ 
                    "Motorcycles",
                  . %in% c("E-Bike", "BICYCLE", "Bike", "E-Scooter") ~ 
                    "Non-Motor Vehicle",
                  . %in% c("FORK", "unk", "UNKNOWN", "OTHER") ~ "Other/Unknown",
                  TRUE ~ "Other/Unknown"
          ))) |>
      rename(vtype1 = vehicle_type_code_1)

crashes <- crashes %>%
  mutate(across(starts_with("vehicle_type_code_2"),
                ~ case_when(
                  . %in% c("2 dr sedan", "3-Door", "4 dr sedan", "4dsd",
                        "Convertible", "Sedan", "SEDONA",
                        "Station Wagon/Sport Utility Vehicle",
                        "SPORT UTILITY / STATION WAGON", "LIMO",
                        "LIVERY VEHICLE") ~ "Passenger vehicles",
                  . %in% c("AMBU", "AMBUL", "Ambulance", "AMBULANCE",
                        "AMBULETTE", "Armored Truck", "Beverage Truck",
                        "Box Truck", "Bulk Agriculture", "Bus", "BUS",
                        "Carry All", "Chassis Cab", "DELIV", "DELV",
                        "Dump", "FDNY", "Fire", "FIRE TRUCK", "FIRETRUCK",
                        "Flat Bed", "Flat Rack", "Garbage or Refuse",
                        "Hopper", "SMALL COM VEH(4 TIRES)",
                        "LARGE COM VEH(6 OR MORE TIRES)",
                        "Mack Truck",
                        "PICK-UP TRUCK", "PK", "Refrigerated Van", "schoo",
                        "SCHOO", "Snow Plow", "Tow Truck",
                        "Tow Truck / Wrecker", "TRACT",
                        "Tractor Truck Diesel", "Tractor Truck Gasoline",
                        "TRAIL", "TRAILER", "TRALI", "TRUCK",
                        "USPS", "UTIL", "Van", "VAN",
                        "VAN TRUCK") ~ "Commercial vehicles",
                  . %in% c("MOTOR SCOO", "Motorcycle",
                        "MOTORCYCLE", "Motorscooter", "SCOOTER", "Moped") ~ 
                    "Motorcycles",
                  . %in% c("E-Bike", "BICYCLE", "Bike", "E-Scooter") ~ 
                    "Non-Motor Vehicle",
                  . %in% c("FORK", "unk", "UNKNOWN", "OTHER") ~ "Other/Unknown",
                  TRUE ~ "None"
          ))) |>
      rename(vtype2 = vehicle_type_code_2)

# boolean variables for vehicle type and MVC causes
crashes <- crashes |>
  mutate(
    involved_motorcycle = if_else(
      vtype1 == "Motorcycles" | vtype2 == "Motorcycles", T, F
    ),
    involved_non_motor = if_else(
      vtype1 == "Non-Motor Vehicle" | vtype2 == "Non-Motor Vehicle", T, F
    ),
    was_impaired = if_else(
      factor1 == "Impairment/Distraction/Fatigue" |
        factor2 == "Impairment/Distraction/Fatigue", T, F
    ),
    failed_to_obey = if_else(
      factor1 == "Failure to Obey Traffic Signs/Signals/Rules" |
        factor2 == "Failure to Obey Traffic Signs/Signals/Rules", T, F
    ),
    mech_failures = if_else(
      factor1 == "Performance-unrelated Technical/Mechanical Factors" |
        factor2 == "Performance-unrelated Technical/Mechanical Factors", T, F
    ),
    misc_cause = if_else(
      factor1 == "Other/Unknown" | factor2 == "Other/Unknown", T, F
    )
  )

# make `factor1`, `factor2`, `vtype1`, and `vtype2` factors 
# rather than characters
crashes <- crashes %>%
  mutate(factor1 = as.factor(factor1),
         factor2 = as.factor(factor2),
         vtype1 = as.factor(vtype1),
         vtype2 = as.factor(vtype2))
```

## Exploratory Analysis

The following figure demonstrates the large difference in casualty rate between 
MVCs that did and did not involve motorcycles or non-motor vehicles. Similar
bivariate analyses showed much smaller differences in casualty rates among 
different causal factors (see appendix).

```{r}
#| label: viz-prop-casualties-by-vtype1-and-vtype2
#| fig-width: 7
#| fig-height: 3

crashes %>%
  filter(vtype1 != "Other/Unknown") %>%
  ggplot(aes(x = vtype1, fill = has_casualty)) +
  geom_bar(position = "fill", alpha = 0.5, color = "black") + 
  scale_fill_manual(values = c("lightgray", "red"), labels = c("No", "Yes")) +
  theme_classic() +
  annotate("text", x = 1, y = 1.05, label = "773 MVCs", color = "red",
           alpha = 0.7, fontface = "bold") +
  annotate("text", x = 2, y = 1.05, label = "79 MVCs", color = "red",
           alpha = 0.7, fontface = "bold") +
  annotate("text", x = 3, y = 1.05, label = "73 MVCs", color = "red",
           alpha = 0.7, fontface = "bold") +
  annotate("text", x = 4, y = 1.05, label = "5557 MVCs", color = "red",
           alpha = 0.7, fontface = "bold") +
  labs(
    x = NULL, 
    y = "Casualty Rate", 
    fill = "Casualty", 
    title = "Motorcycles and non-motor vehicles have the highest casualty rate",
    subtitle = "Casualty rate of MVCs in New York City by vehicles involved"
  )
```

# Methodology

We fit a multivariable logistic regression model examining where there are 
associations between odds of casualty occurrence and peri-accident 
characteristics. Our predictors were chosen a priori using our original analyses 
(shown above and in the appendix) and based on associative factors identified in 
prior published research in the field (Mohamed et al., 2013; Zou et al., 2017). 

Independence is reasonable because exactly one observation is made per accident,
regardless of number of vehicles involved. Thus, we assume that accidents are
independent from each other and that one accident cannot impact others. 
Linearity, as assessed via empirical logit plots, was also reasonable 
(see appendix).

All statistical analyses were performed at an a priori 0.05 level of 
significance.

# Results

Our model found seven statistically significant $(p<0.05)$ predictors of
casualties in MVCs. As expected from our exploratory analysis, MVCs have a much
higher chance of casualty if motorcycles $(p<0.001)$ or non-motor 
vehicles $(p<0.001)$ were involved when compared to a baseline of MVCs which
involved neither, with odds ratios of 10.2 and 18.9 respectively. 

Further, evening- and night-time MVC's $(p<0.001)$ showed an increased casualty
rate, with odds ratios of 1.45 and 1.73 respectively, compared to a baseline
of morning accidents. The most significant contributing factors were failure to 
obey traffic rules $(p<0.001)$, impairment $(p<0.001)$, and unknown or 
miscellaneous $(p<0.001)$, with odds ratios of 2.04, 1.36, and 1.46 
respectively, when compared to a baseline contributing factor of 
aggressive/reckless driving.

Table 1 displays selected odds ratios and associated 95\% confidence intervals
from our model. Full model results, including estimates from all predictors and
a complete description of the baseline MVC, are available in the appendix.

```{r}
#| label: final-model-output
#| tbl-cap: Significant predictors (p < 0.05)
finalmodel <- glm(has_casualty ~ involved_motorcycle + involved_non_motor +
                    time_day + weekend_weekday + yday + failed_to_obey + 
                    was_impaired + mech_failures + misc_cause, 
                  data = crashes, family = "binomial")

finalmodel_tidy <- tidy(finalmodel)

# add an odds column to final model output by exponentiating slope estimates
finalmodel_tidy <- finalmodel_tidy %>%
  mutate(odds = exp(estimate))

# output table manually
pred_names <- c("Involved motorcycle", "Involved non-motor vehicle",
               "MVC during evening", "MVC at night", 
               "Failed to obey traffic rules", "Impaired driver",
               "Miscellaneous/unknown cause")

ci <- confint(finalmodel) |> 
  exp() |> 
  round(2) |> 
  as_tibble() |>
  slice(2,3,5,6,9,10,12) |>
  mutate(joined = paste("[", `2.5 %`, ", ", `97.5 %`, "]", sep = "")) |>
  select(joined) |>
  pull()

tab_data <- finalmodel_tidy |>
  filter(p.value <= 0.05) |>
  select(term, odds, p.value) |>
  mutate(
    p.value = format(p.value, scientific = T, digits = 2),
    odds = round(odds, 2)
  ) |>
  slice_tail(n = 7) |>
  rename(Predictor = term, `Odds ratio` = odds, `P-value` = p.value) |>
  mutate(
    Predictor = pred_names,
    `95% CI` = ci
  )


table <- flextable(tab_data)
table <- width(table, width = 1.3)
table <- align(table, align = "right", part = "all")
table <- width(table, j = 1, width = 2.5)
table <- bold(table, part = "header")
table
```

# Discussion

Whether or not an MVC involved a cyclist or motorcycle rider has by far the 
largest odds of casualties, with odds ratios of 18.89 and 10.25 respectively,
compared to the next highest odds ratio (from failure to obey traffic rules)
being 2.04. However, should policy makers be willing to make only a single 
change to try to stem the flow of MVC casualties, improving conditions for 
bikers may not be the most efficient choice. While their casualty rate is the 
highest, in absolute number, non-motor vehicles and motorcycles combine to make 
up less than 200 of the 10,000 MVCs in our sampling.

To save the highest absolute number of casualties, improving evening- and 
night-time conditions is a better option, as the total number of casualties
across times of day is much more uniform than across vehicle type. 

While we were unable to investigate the specific causes of the increase in 
casualty rate during the later hours of the day, future research that includes 
light conditions from sunlight or streetlights, weather data, impairment rates, 
and other possible causes could further narrow in on a policy recommendation for
city leaders. Additional research is also needed to investigate whether or not
working to increases rates of cycling (thus having fewer cars to cause 
accidents) would be a net increase or decrease in absolute casualty count, 
especially given such high casualty rates in the accidents that *do* occur
involving cyclists and motorcyclists.

While further research and additional data are needed, we hope that the insights 
that our model is able to predict can at least spark a conversation about 
potential policy decisions for urban planners in NYC and similar American 
cities.

{{< pagebreak >}}

# Appendix
## Data Dictionary

`has_casualty` (response variable): A boolean variable that is `True` when there 
was at least one injury or death in an MVC.

`involved_motorcycle`: A boolean variable that is `True` when one or more of the
vehicles involved was a motorcycle, motor scooter, or moped (generalized to 
"motorcycle" in this report).

`involved_non_motor`: A boolean variable that is `True` when one of the vehicles
involved was a bicycle or scooter (electric or analog).

`time_day`: A categorical variable that is `morning` if the MVC occurred between
5:00 am and 11:59 am, `afternoon` if between 12:00 pm and 4:59 pm, `evening` if
between 5:00 pm and 8:59 pm, and `night` if between 9:00 pm and 4:49 am.

`weekend_weekday`: A categorical variable that is `Weekend` if an MVC occurred
on Saturday or Sunday, and `Weekday` otherwise.

`yday`: A numeric variable representing the cumulative day of the year (1-365)
on which an MVC occurred.

`failed_to_obey`: A boolean variable that is `True` if one of the listed causes
was failure to obey traffic signs/signals/rules.

`was_impaired`: A boolean variable that is `True` if one of the listed causes
was impairment, fatigue, or fatigue.

`mech_failures`: A boolean variable that is `True` if one of the listed causes
was performance-unrelated mechanical failures of one or more vehicles.

`misc_cause`: A boolean that is `True` if the cause of an MVC is unknown or does
not fall into any other category.

## Exploratory Data Analysis

```{r}
#| label: viz-time-density-with-weekend_weekday
#| fig-width: 7
#| fig-height: 3
crashes |>
  filter(num_casualties > 0) |>
  mutate(color = if_else(weekend_weekday == "Weekend", 
         "red", "gray")) |>
  ggplot(aes(x = crash_time, group = weekend_weekday)) +
  geom_density(color = "lightgray") +
  geom_density(
    data = filter(crashes, num_casualties > 0, 
                  weekend_weekday == "Weekend"), 
    color = "red"
  ) +
  geom_vline(xintercept = hm("9:00"), linetype = 3) +
  geom_vline(xintercept = hm("17:00"), linetype = 3) +
  annotate("text", label = "9:00 am", color = "gray", angle = 90,
           x = hm("10:00"), y = 5e-06, hjust = "right", vjust = "bottom") +
  annotate("text", label = "5:00 pm", color = "gray", angle = 90,
           x = hm("16:30"), y = 5e-06, hjust = "right", vjust = "bottom") +
  annotate("text", color = "red", label = "Weekend", x = hm("00:20"), 
           hjust = "left", fontface = "bold", y = 8e-06) +
  annotate("text", color = "gray", label = "Weekday", x = hm("00:20"), 
           hjust = "left", fontface = "bold", y = 1e-06) +
  scale_color_identity() +
  theme_classic() +
  labs(
    x = "Time of crash",
    y = "Density",
    title = "Weekend casualties are disproportionately in the early morning",
    subtitle = paste("Time density of NYC car accidents with injuries or",
                     "fatalities by Weekend or Weekday"),
    color = "Weekend or Weekday"
  )
```


```{r}
#| label: viz-zip-code-box
#| fig-height: 3
#| fig-width: 7
zip_crash <- crashes |>
  group_by(zip_code) |>
  summarise(count = n(), casualties = sum(has_casualty)) |>
  mutate(prop = casualties / count)

zip_med <- median(zip_crash$prop)

zip_crash |>
  ggplot(aes(x = prop)) +
  geom_vline(xintercept = zip_med, linetype = 3) +
  annotate("text", label = "Median = 0.186", x = 0.2, y = 0.8, color = "gray",
           fontface = "bold", hjust = "left") +
  geom_boxplot(varwidth = T, fill = "red", alpha = 0.5) +
  ylim(-1,1) +
  labs(
    x = "Proportion of accidents with casualties",
    y = NULL,
    title = "Zip Code casualty rates are approximately normal",
    subtitle = "Distribution of NYC car accident casualty rates by zip code"
  ) +
  theme_classic() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank()
  )
```


```{r}
#| label: viz-prop-casualties-by-factor1-and-factor
#| fig-width: 7
#| fig-height: 3
# factor1
crashes %>%
  filter(factor1 != "Other/Unknown") %>%
  ggplot(aes(x = factor1, fill = has_casualty)) +
  geom_bar(position = "fill", color = "black", alpha = 0.5) + 
  scale_fill_manual(values = c("lightgray", "red"), labels = c("No", "Yes")) +
  scale_x_discrete(
    labels = c("Aggressive Driving", "Traffic Violations", 
               "Impairment/Distraction", "Technical Failure")) +
  theme_classic() +
  labs(
    x = NULL, 
    y = "Casualty Rate", 
    fill = "Casualty", 
    title = "No contributing factor is disproportionally casualty-prone", 
    subtitle = "MVCs with casualties based on cause of accident"
  )
```

## 3. Methodology

#### a) Likelihood-ratio test between models considered

The following output shows that `Model 2` is superior $(p<0.001)$, 
and was therefore used in our final analysis.

```{r}
#| label: build-models and likelihood ratio test
# model 1: base model
model_1 <- glm(has_casualty ~ involved_motorcycle + involved_non_motor + 
                 time_day + weekend_weekday + yday, 
              data = crashes, family = "binomial")

# model 2: add contributing factors
model_2 <- glm(has_casualty ~ involved_motorcycle + involved_non_motor +
                 time_day + weekend_weekday + yday + failed_to_obey + 
                 was_impaired + mech_failures + misc_cause, 
              data = crashes, family = "binomial")

# model 3: add `weekend_weekday*time_day` interaction
model_3 <- glm(has_casualty ~ involved_motorcycle + involved_non_motor +
                 time_day + weekend_weekday + yday + failed_to_obey +
                 was_impaired + mech_failures + misc_cause +
                 weekend_weekday*time_day, 
              data = crashes, family = "binomial")

# Perform likelihood ratio test
options(show.models = F)
anova <- anova(model_1, model_2, model_3, test = "Chisq")
anova[-1]
```


#### b) Numerical Day of the Year vs. Log-odds of Casualty Plot

```{r}
#| label: emplogit
emplogitplot1(has_casualty ~ (yday), data = crashes, ngroups = "all")
```


{{< pagebreak >}}

## 4. Final model

The following table is the complete model used in our analysis. Variable names
are raw, as seen in our code. These results are compared to a baseline MVC in 
which no non-motor vehicle or motorcycle was involved, the cause of the MVC
was aggressive/reckless driving, and the MVC occurred on January 1st and a
weekday.

```{r}
#| label: full-model-output
#| tbl-cap: Full model output
terms <- c("Baseline", "Involved Motorcycle", "Involved non-motor vehicle",
           "Afternoon", "Evening", "Night", "Weekend", "Day of year",
           "Failed to obey", "Impaired", "Mech. Failures", 
           "Misc./Unknown cause")
finalmodel_tidy |>
  mutate(
    estimate = format(estimate, scientific = T, digits = 3),
    std.error = format(std.error, scientific = T, digits = 3),
    p.value = format(p.value, scientific = T, digits = 3),
    odds = format(odds, scientific = T, digits = 3),
    term = terms
  ) |>
  select(-std.error, -statistic) |>
  flextable() |>
  align(align = "right", part = "all") |>
  bold(part = "header") |>
  width(width = 1.3) |>
  width(j = 1, width = 2.5)
```


{{< pagebreak >}}

# 5. Works Cited

Department of Health. All Injuries in New York State. (n.d.). Retrieved May 4, 
2023, from https://www.health.ny.gov/statistics/prevention/injury_prevention/all_injury.htm

Mohamed, M. G., Saunier, N., Miranda-Moreno, L. F., &amp; Ukkusuri, S. V. 
(2013). A clustering regression approach: A comprehensive injury severity 
analysis of pedestrian–vehicle crashes in New York, US and Montreal, Canada. 
Safety Science, 54, 27–37. https://doi.org/10.1016/j.ssci.2012.11.001 

Zou, W., Wang, X., &amp; Zhang, D. (2017). Truck crash severity in New York 
City: An investigation of the spatial and the time of day effects. Accident 
Analysis &amp;amp; Prevention, 99, 249–261. 
https://doi.org/10.1016/j.aap.2016.11.024 